---
title: "Sub-commands"
author: "Zuguang Gu (z.gu@dkfz.de)"
date: '`r Sys.Date()`'
output:
  html_document:
    fig_caption: true
    toc: true
    toc_depth: 3
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Sub-commands}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---------------------------------------------------------------------

```{r, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
    tidy  = FALSE,
    comment = NA,
    fig.align = "center")
library(GetoptLong)
```


## A quick example

The following code gives you a quick example of how the sub commands are set. In the function `subCommands()`, the input is in form
of a vector, where the elements are grouped by every three. The three elements in every group are 

1. name of the sub command,
2. path of the corresponding R script,
3. description of the corresponding sub command.

The following code is in the script called `main.R` which contains dispatch of the sub commands.

```{r main, eval = FALSE}
library(GetoptLong)

subCommands(
	"sub1", "sub1.R", 
	        "This is the description of sub command 1, which will be a long long long text.",
	"sub2", "sub2.R",
	        "This is the description of sub command 2, which will be a long long long text."
)
```

Code in `sub1.R`:

```{r sub1, eval = FALSE}
library(GetoptLong)

GetoptLong("foo=i", "Set value for foo.")

qqcat("foo = @{foo}\n")
```

And code in `sub2.R`.

```{r sub2, eval = FALSE}
library(GetoptLong)

GetoptLong("bar=i", "Set value for bar")

qqcat("bar = @{bar}\n")
```

Calling `main.R` with no argument prints the list of sub commands it supports:

```
Rscript main.R
```

```{r, echo = FALSE, eval = TRUE, results = "asis"}
chunks <- knitr:::knit_code$get()
writeLines(chunks[["main"]], con = "main.R")
writeLines(chunks[["sub1"]], con = "sub1.R")
writeLines(chunks[["sub2"]], con = "sub2.R")
cat("```\n")
cat(system("Rscript main.R", intern = TRUE), sep = "\n")
cat("```\n")
```

When sub command is called, the corresponding R script is invoked. Here all command-line arguments after the sub command `sub1` go
to the correspond R script `sub1.R`.

```
Rscript main.R sub1 --help
```

```{r, echo = FALSE, eval = TRUE, results = "asis"}
cat("```\n")
cat(system("Rscript main.R sub1 --help", intern = TRUE), sep = "\n")
cat("```\n")
```

Internally, it actually executes `Rscript sub1.R --help`.

Options in `sub1.R` can be directly set after the sub command:

```
Rscript main.R sub1 --foo 10
```

```{r, echo = FALSE, eval = TRUE, results = "asis"}
cat("```\n")
cat(system("Rscript main.R sub1 --foo 10", intern = TRUE), sep = "\n")
cat("```\n")
```


## The specification

The specification for sub commands can be set as a vector. The length should be multiples of either 2 or 3.

In the following code, the input of `subCommands()` will be internally put into a three-column matrix filled by rows,
where the first column contains sub commands, the second column contains paths of corresponding R scripts, and the third
column contains descriptions for sub commands.

```{r, eval = FALSE}
subCommands(
	"sub1", "path/sub1.R", "description 1",
	"sub2", "path/sub2.R", "description 2"
)
```

Or in another way:

```{r, eval = FALSE}
subCommands(
	"path/sub1.R", "description 1",
	"path/sub2.R", "description 2"
)
```

In this case, the sub commands are taken from the basename of the R script, also removing the suffix names (`.R`).

Whether the arguments are grouped by every 2 elements or 3 elements is automatically detected in `subCommands()`.

## Commands in sections

Sometimes are want to group commands and put then into sections. This can be done by adding a section tag `"-----"` as shown in the following example.

```{r section, eval = FALSE}
subCommands(
	"----", "----", "This is section1",
	"sub1", "sub1.R", 
	      "This is the description of sub command 1, which will be a long long long text.",

	"----", "----", "This is section2",
	"sub2", "sub2.R",
	      "This is the description of sub command 2, which will be a long long long text."
)
```


```{r, echo = FALSE, eval = TRUE, results = "asis"}
chunks <- knitr:::knit_code$get()
writeLines(c("library(GetoptLong)", chunks[["section"]]), con = "main.R")

cat("```\n")
cat(system("Rscript main.R", intern = TRUE), sep = "\n")
cat("```\n")
```

## Setting help head and foot

`help_head` adds text before the help message and `help_foot` adds text after the help message.


```{r head-foot, eval = FALSE}
subCommands(
	help_head = "This is the head of the help message.",

	"sub1", "sub1.R", 
	        "This is the description of sub command 1, which will be a long long long text.",
	"sub2", "sub2.R",
	        "This is the description of sub command 2, which will be a long long long text.",

	help_foot = "This is the foot of the help message."
)
```

```{r, echo = FALSE, eval = TRUE, results = "asis"}
chunks <- knitr:::knit_code$get()
writeLines(c("library(GetoptLong)", chunks[["head-foot"]]), con = "main.R")

cat("```\n")
cat(system("Rscript main.R", intern = TRUE), sep = "\n")
cat("```\n")
```

```{r, echo = FALSE, results = "hide"}
file.remove("main.R")
file.remove("sub1.R")
file.remove("sub2.R")
```

<script>
var elements = document.getElementsByTagName("pre");
for(var i = 0; i < elements.length; i ++ ) {
    if(!elements[i].hasAttribute("class")) {
        code = elements[i].children[0];
        code.style.color = "white";
        elements[i].style.background = "hsl(70, 0%, 15%)";
    }
}
</script>
